On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-22 paper/paper3d.tex
starting another revision
re-writing intro

The standard metric for the (inverse) strength of disk self-gravity is
the \emph{Toomre parameter}, 
\begin{align}\label{toomreQ_criterion}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}  
\end{align}
\citep{toomre64}. Here, $c_s$ is the sound-speed,
$\kappa$ is the epicyclic frequency (which equals the rotation
frequency $\Omega$ in a Keplerian disk), $\Sigma$ is the surface
density and $G$ is the gravitational constant. The Toomre $Q$ is a
measure of the destabilizing effect of self-gravity ($G\Sigma$) against 
the stabilzing effect of rotation ($\kappa$) and pressure 
($c_s$). This is evident from the dispersion relation for local,
axisymmetric waves in a two-dimensional (2D, razor-thin), 
inviscid and adiabatic (including isothermal) disk, 
\begin{align} 
  s^2 = \kappa^2 + c_s^2k^2 - 2\pi G\Sigma|k|, 
\end{align}
relating the growth rate $s$ and the radial wavenumber $k$. When
$Q\leq 1$ the disk is formally unstable.

%assess GI in general

The Toomre parameter is widely applied to assess GI in more general 
situtations (cf. the ideal disk models from which $Q$ is derived
from). This includes realistic disk structures 
that account for non-ideal physics not 
explicitly reflected by $Q$ \citep[e.g.][]{kimura12}.  It is, however,
not obvious that such additions leave the stability condition
unaltered.

Indeed, we show that when cooling and viscosity are  
included, as expected in realistic disks, GI cannot be assessed from a
single `Toomre-like' condition.      
%disk structure may be computed with sophisticated physics but
%stability condition still based on ideal disk models. 
%three main ingredients - gravity, pressure, rotation 
%a standard way to apply this condition is to compute the disk
%structure  based on realistic physics 
The goal of this work is to generalize the analytic treatment of disk 
GI to allow a self-consistent discussion of GI in realistic accretion 
disk models. 
%Specifically, we will examine the effect of cooling and  
%viscosity on GI in PPDs.

%Their relevance to PPDs are described below.

%stuff aboout gravito turbulence 
%Th
%we focus on cooling and viscosity but also do 3D 
%but real disks have cooling and viscosity/turbulence
%(e.g. gravito-turbulent disks)

\subsection{Cooling-driven gravitational instability}
A net thermal loss\footnote{Isothermal disks implicitly assume 
  heating and cooling processes are exactly balanced at all
  times, so these do not have cooling in the current context.} (hereafter `cooling')
reduces pressure (thermal) support against self-gravity. The cooling time 
$\tcool$ is the time-scale over which the disk temperature $T$ is relaxed
to some pre-defined value $T_0$, which may be zero. That is, $\p_t T =
-(T-T_0)/\tcool$. It is often written as 
\begin{align} 
  \tcool = \beta \Omega^{-1},
\end{align}
where $\beta$ is the corresponding dimensionless cooling time. This
type of cooling function, first applied by \cite{gammie01}, is a
simple parameterization that allows the full range of thermodynamic
response to be explored. 
{\bf physical origin of cooling in PPDs}

The Toomre $Q$ parameter does not include cooling, and thus cannot be
formally applied to assess GI in realistic disks. Indeed, there exists
a separate, \emph{empirical} cooling time criterion which governs
stability against disk fragmentation (see below), but this is not well
understood.

We shall quantify how cooling removes pressure
support to enable GI even when $Q>1$. 
We will find that cooling in fact changes the 
qualitative nature of GI, in that its mere presence can
formally lead to instability.

@@

\subsection{Relevance to gravito-turbulent disk fragmentation}%and ring
                                %formation 
%can the continous formation/destruction of clumps (as seen in sims)
%be interpreted as `overstability'? 
%for overstable modes over-density oscilates 
The two effects described above are immediately relevant to
`gravito-turbulence'. This is a steady disk state in which cooling is
balanced by turbulent heating, itself caused by GI. In this case 
the $\alpha$ and $\beta$ parameters are inversely related
\citep{gammie01}.

>>>>

The standard metric for the (inverse) strength of disk self-gravity is
the \emph{Toomre parameter}, 
\begin{align}\label{toomreQ_criterion}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}  
\end{align}
\citep{toomre64}. Here, $c_s$ is the sound-speed,
$\kappa$ is the epicyclic frequency (which equals the rotation
frequency $\Omega$ in a Keplerian disk), $\Sigma$ is the surface
density and $G$ is the gravitational constant. The Toomre $Q$ is a
measure of the destabilizing effect of self-gravity ($G\Sigma$) against 
the stabilzing effect of rotation ($\kappa$) and pressure 
($c_s$). This is evident from the dispersion relation, 
\begin{align}\label{classic_dr}
  s^2 =  2\pi G\Sigma|k| - \kappa^2 - c_s^2k^2,   
\end{align}
which relates the growth rate $s$ and radial wavenumber $k$ for local,
axisymmetric waves in a two-dimensional (2D, razor-thin), inviscid and
adiabatic (including isothermal) disk. Such disturbances are formally
unstable when the Toomre condition, $Q<1$, is satisfied.
Non-axisymmetric modes can develop for  larger but still order-unity
values of $Q$ . {\bf ref} 
%strictly speaking non axi *local* modes are stable -> only transient
%growth
%true non-axi instability needs global disk  model 
%non-axi GI at slightly larger Q, but still characterized by Q

%qualitative intro to gravito-turbulence 
%role of cooling and resulting viscosity

It is important to remember that the dispersion relation
Eq. \ref{classic_dr}  is derived from \emph{idealized} conditions: 
the disk is laminar and does not experience any net thermal
losses\footnote{Isothermal disks implicitly assume  
  heating and thermal losses are exactly balanced at all 
  times, so these do not cool in the current context.} 
(hereafter `cooling').  
However, real disks can cool, for example,
due to radiative losses. Accretion disks may also be turbulent, for
which its dynamic and thermodynamic effect is often modeled through an
effective viscosity \citep{shakura73,lin87,rafikov15}. {\bf 
  more refs here? non-sg refs} 
%in terms of AM transport and heating 
%assess GI in general

Despite this, the Toomre parameter, and its associated
dispersion relation, are still widely applied to realistic disk models
with viscosity and cooling \citep[e.g.][]{cossins09, kimura12}, 
even though Eq. \ref{toomreQ_criterion}---\ref{classic_dr} were 
derived from inviscid disks without cooling.

However, as we emphasize in this work, the inclusion of non-ideal
physics, such as cooling and viscosity, modifies the dispersion
relation, Eq. \ref{classic_dr}, and hence the condition for GI. This 
means that the Toomre condition alone is insufficient to assess GI
in realistic disks. The goal of this work is to generalize the
analytic treatment of disk GI to allow a self-consistent discussion of
GI in realistic accretion disk models. Specifically, we consider the non-ideal
effects of viscosity and cooling on GI.

%Specifically, we will examine the effect of cooling and  
%viscosity on GI in PPDs.

\subsection{Cooling-driven gravitational instability}
Cooling reduces pressure (thermal) support against self-gravity. The cooling time 
$\tcool$ is the time-scale over which the disk temperature $T$ is relaxed
to some pre-defined value $T_0$, which may be zero. That is, $\p_t T =
-(T-T_0)/\tcool$. It is often written as 
\begin{align} 
  \tcool = \beta \Omega^{-1},
\end{align}
where $\beta$ is the corresponding dimensionless cooling time. This
type of cooling function, first applied by \cite{gammie01}, is a
simple parameterization that allows the full range of thermodynamic
response to be explored. 
{\bf physical origin of cooling in PPDs}

%The Toomre $Q$ parameter does not include cooling, and thus cannot be
%formally applied to assess GI in realistic disks. Indeed, there exists
%a separate, \emph{empirical} cooling time criterion which governs
%stability against disk fragmentation (see below), but this is not well
%understood.

We shall quantify how cooling removes pressure support to enable GI
even when $Q>1$.

We will find that cooling in fact changes the 
qualitative nature of GI, in that its mere presence can
formally lead to instability.

@@

\subsection{Relevance to gravito-turbulent disk fragmentation}

% Then the disk temperature will be lowered 
% until $Q\sim 1$, whence non-axisymmetric modes grow 
% and heats the disk through dissipation of spiral shocks. This may 
% lead to a quasi-steady turbulent state with $Q\sim 1$ in
% which cooling is balanced by shock heating to maintain thermal
% equilibrium \citep{gammie01}.

The two effects described above are immediately relevant to
`gravito-turbulence'. This is a steady disk state in which cooling is
balanced by turbulent heating, itself caused by GI. In this case 
the $\alpha$ and $\beta$ parameters are inversely related
\citep{gammie01}.
